```{r}
rm(list = ls(all = TRUE))
library(INLA)
```

```{r}
housing<-read.csv("housing.csv")
#removing rows with NA's, there are only a few of these 
housing=housing[complete.cases(housing), ]
#creating a new covariate 
housing$average_bed_rooms=housing$total_bedrooms/housing$households
head(housing)

#We split the original dataset into two parts, training and test
housing.training<-housing[seq(from=1,to=nrow(housing),by=2), ] 
housing.test<-housing[seq(from=2,to=nrow(housing),by=2), ]
```

**Q1)[10 marks]**					

**Fit a Bayesian Linear regression model in INLA (with Gaussian likelihood) using the housing.training dataset such that the response variable is the log(median_house_value), and the covariates in the model are as follows:** **longitude, latitude, housing_median_age, log(median_income), ocean_proximity, average_bed_rooms.**			

**Use scaled versions of the non-categorical covariates in your model.**

```{r}
names(housing)
```

```{r}
# Install and load necessary packages
install.packages("data.table")
library(data.table)

# Prepare the data
housing.training$log_median_income <- log(housing.training$median_income)
housing.test$log_median_income <- log(housing.test$median_income)

# Scale the non-categorical covariates
housing_scaled.training <- scale(housing.training[, c("longitude", "latitude","housing_median_age", "log_median_income", "average_bed_rooms")])
housing_scaled.test <- scale(housing.test[, c("longitude", "latitude", "housing_median_age","log_median_income", "average_bed_rooms")])

# Transform the response variable
housing.training$log_median_house_value <- log(housing.training$median_house_value)
housing.test$log_median_house_value <- log(housing.test$median_house_value)

# Combine scaled covariates and transformed response variable
housing_scaled.training <- cbind(housing_scaled.training, housing.training$log_median_house_value, as.factor(housing.training$ocean_proximity))
housing_scaled.test <- cbind(housing_scaled.test, housing.test$log_median_house_value, as.factor(housing.test$ocean_proximity))

# Convert to data.table
housing_dt <- data.table(housing_scaled)

# Define the model formula
formula <- log_median_house_value ~ longitude + latitude + housing_median_age + log_median_income + average_bed_rooms + ocean_proximity

# Fit the model using INLA
model <- inla(formula, data = housing_scaled.training, control.predictor = list(compute = FALSE))

# Print out the model summary
summary(model)

```
